Change in Mobility and Site Occupation during the Late Pleistocene in Korea

### 2019-11-22 analysis/paper/paper.Rmd
Added a bit for the discussion. I didn't finish yet but just commiting/pushing before I go home...

Radiocarbon dates provide a proxy for change in human population levels. Summed probability distributions (SPD) of radiocarbon dates are often used to understand temporal trends in ancient human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed]. Without there must be unavoidable limitations for the validity of SPD such as the lack of excavated sites in certain period, this method is still helpful to be used as a proxy for population fluctuation. Timpson et al. demonstrate its usefulness by testing data sets in 12 European regions and assert that this SPD is extremely robust to deal with small sample size [@timpson2014reconstructing]. We use 171 radiocarbon dates from 22 sites to search for population trends during the Late Pleistocene in Korea based on the premise that population change or pressure could be another catalyst to improve or innovate technology.

@@

The new technology and occupation in high elevation
The new technology appeared around 40-35ka in the Korean Peninsula includes stemmed points and following blade industry. The stemmed points brought new paradigms in stone artifact assemblages. For instance, the concept of composite tool contributed to increase efficiency of maintaining the toolkit by easily replacing the damaged part [@kuhn2015artifacts]. Combining with other materials increases utility or functionality of the tool by adding extra weight and providing a shaft, which makes it earlier to grip. Selective use of raw material might impact on hunter-gathers’ mobility and occupation pattern depending on availability of the specific raw materials such as shale and rhyolite. In addition, this research shows that carrying the points enabled hunter-gatherers to occupy higher elevation compared to the sites with no stemmed points. In other words, the new technology might play an important role to survive in environments with more risks.

Population change
Interpreted in HBE, we assume there might be close connection between environmental change and the appearance of new technology based on the prediction that the new invention was the result of risk. The research suggests that the environmental pressure was related to population change. We cannot simply predicate that the increased population around 35ka was due to dispersal of Homo Sapiens since there is no reliable human remains in South Korea. However, the migration of Homo sapiens was not impossible scenario at the same time. Bae (2010) argues that the emergence of stemmed points and blade industry came from outside of the Korean peninsula. He proposes that the new technology was the result of the continuous influx of modern human populations from Siberia while the earlier coarse flake tradition ----

>>>>

Radiocarbon dates provide a proxy for change in human population levels. Summed probability distributions (SPD) of radiocarbon dates are often used to understand temporal trends in ancient human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed]. Even though there must be unavoidable limitations for the validity of SPD such as the lack of excavated sites in certain period, this method is still helpful to be used as a proxy for population fluctuation. Timpson et al. demonstrate its usefulness by testing data sets in 12 European regions and assert that this SPD is extremely robust to deal with small sample size [@timpson2014reconstructing]. We use 171 radiocarbon dates from 22 sites to search for population trends during the Late Pleistocene in Korea based on the premise that population change or pressure could be another catalyst to improve or innovate technology.

@@

1. The new technology and occupation in high elevation
The new technology appeared around 40-35ka in the Korean Peninsula includes stemmed points and following blade industry. The stemmed points brought new paradigms in stone artifact assemblages. For example, projectile points are only a part of composit artifacts and this specific characteristics could contribute to increase efficiency of maintaining the toolkit by easily replacing the damaged part [@kuhn2015artifacts; @cardillo2010some]. Combining with other materials increases utility or functionality of the tool by adding extra weight and providing a shaft, which makes it earlier to grip. Selective use of raw material might impact on hunter-gathers’ mobility and occupation pattern depending on availability of the specific raw materials such as shale and rhyolite. In addition, this research shows that carrying the points enabled hunter-gatherers to occupy higher elevation compared to the sites with no stemmed points. Therefore, we can assume that the technology would play an important role to survive in more risky environments.

2. Population change
Interpreted in HBE, we expected there might be close connection between environmental change and the appearance of new technology based on the prediction that the new invention was the result of risk. The research suggests that the environmental pressure coule be related to population change. We cannot simply predicate that the increased population around 35ka was due to dispersal of Homo Sapiens since there is no reliable human remains in South Korea. However, the migration of Homo sapiens was not impossible scenario at the same time. In fact, one model explains that the emergence of stemmed points and blade industry came from outside of the Korean peninsula and the new technology was the result of the continuous influx of modern human populations from Siberia while the earlier coarse flake trandition was come from southern China [@Bae_2010]. Since this migration model barely explains about preexisted stone artifact assemblages and people in the Korean Peninsula, it is hard to accept his argument without knowing about the native people. In short, we assume that the stemmed points appeared related to the increased population and it can be related to the modern human dispersal or indigenously increased occupants.

3. Expedient vs curated
To monitor moibility strategies and site occupation pattern, we examined the artifact density and retouch frequency, based on the premise that highly mobile foragers would have light weighted, highly retouched, replaceable, almost completely made tools and leave small amount of artifacts when they move to the new location (curated). One the other hand, less mobile foragers who tend to stay in one place yield more debitage and debris, produce less efficient tools since they can easily get supplied (expedient) [@clark2019landscapes; @kuhn1994formal; @kuhn2014mousterian; @meignen2006middle]. Therefore, the curated technology indicates sites with special or higly mobile groups with stone artifacts that are thoroughly designed and based on a high level of investment of time and effort in artifact manufacture and maintenance. The expedient technology represents base camp or residence with artifacts that tend to be manufactured, used, and abandoned in the short period of time [@vaquero2018searching;@binford1979organization; @torrence1989time].

During the transitoin period, sites were used in
